### 2022-02-07 paper/paper.tex
proposal: restructure paper and proofread content

\documentclass{report}

@@

@@

@@

@@

	\textsc{Development of a touch typing trainer\\with an emphasis on finger and wrist positions}
	\par\vspace{0.5cm}

	\hrulefill
	\par\vspace{0.25cm}
	A Special Project Proposal Presented to the\\
	Faculty of the Department of Computer Science,\\
	University of the Philippines Cebu

	\par\vspace{0.25cm}
	In Partial Fulfillment\\
	Of the Requirements for the Degree\\
	Bachelor of Science in Computer Science\\
	\par\vspace{0.25cm}
	\hrulefill
	\par\vspace{0.5cm}

@@

	October 2021
	\vfill
	\hspace{0pt}
\end{titlepage}
\newpage

\printacronyms
\newpage

@@

%FIXME - Lots of passive voice
\section{Background of the Study}
There are a lot of educational typing tests available that help people learn
touch typing, including Monkeytype, TypeRacer, and Keybr. These typing tests list
out words that are then typed out. The inputted keys are then compared to check
if the user has typed the expected letter. At the end of the test, the time
taken is calculated, and certain metrics are given. These metrics include words
per minute (WPM) and accuracy \parencite{bartnik2021}.

@@

Correct finger placement is usually taught at the beginning using a diagram,
with each key being associated with a specific finger. For instance, the letter
Q in a QWERTY layout should be hit using the fifth digit of the left hand, and
this is shown by coloring the fifth digit and the key Q with the same color or
by placing the letters directly on the fingers \parencite{dobson2009touch}.

Incorrect finger placement may cause these hand and wrist positions: ulnar
deviation, forearm pronation, and wrist extension \parencite{serina1999}. These
three are hand and wrist positions that are common in all activities, however,
prolonged periods in these positions may cause injuries such as Carpal
tunnel syndrome (CTS) \parencite{toosi2015}

@@

Thus, there is a need for automatically detecting which finger is used during
typing, and for the position of the wrist in relation to the arm. One way to do
this is through finger and hand tracking. One solution for tracking is by using
image processing and machine learning. An example of this is MediaPipe
by~\cite{mediapipe}.

MediaPipe allows for various applications for machine learning in the field of
image processing. This includes, hand tracking, pose estimation, object
detection, and others. Another example of a library that allows for hand and
finger tracking is OpenCV by~\cite{opencv}. This is a tool that simplifies
computer vision and image processing. Machine learning can also be used with
OpenCV.

@@

\subsection{Specific Objectives}
\begin{itemize}
	\item To develop a program that tracks fingers and hand positions while typing
	\item To create a subroutine that ascertains which finger was used to type a key
	\item To detect if incorrect fingers were used to press a key or if the
	      position of the hand in relation to the wrist is problematic
	\item To assess the accuracy and performance of the developed program's ability
	      to perform the previously stated objectives
	\item To develop a user-friendly interface for users to train touch typing
	      using the developed program
	\item To generate key statistics of a user's touch typing performance:
	      \begin{itemize}
		      \item Words per Minute
		      \item Accuracy
		      \item Finger Placement Accuracy
	      \end{itemize}
\end{itemize}

@@

In addition, the keyboard layout will be \ac{ansi}. This layout is described by the
American National Standards Institute \parencite{ansi}. This is the most
common layout in the United States. However, it is also used in numerous
English-speaking countries such as the Philippines, Malaysia, and India.

@@

Capturing of the video to be analyzed by the program would be limited to a
single 480p webcam that is capturing in 24 frames per second. The camera will
be pointed downwards facing the keyboard and the hand. This means that the
vertical angle of the wrist may not be accurate.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{60.jpg}
	\caption{A 60\% keyboard in \ac{ansi} layout. Reprinted from \fullcite{60}}
	\label{fig:60}
	\centering
\end{figure}

\section{Significance of the Research}
This research is beneficial for all users of physical keyboards. These include a
vast majority of the population as there are a lot of professions that heavily
rely on keyboards. Examples include developers, physicians, educators,
accountants. By having better ergonomics while typing, wrist injuries can be
prevented, and typing speed may be increased

@@

This research has a direct impact on people that has hand or wrist injuries that
are caused by poor typing habits. By correcting these poor habits, pain from
these injuries will be lessened, and even be prevented from occurring in the
first place. A specific example of this is by reducing ulnar deviation which
affects the nerve that is indicative of CTS \parencite{toosi2015}.

\chapter{Preliminary Review of Related Literature}

@@

where, $|T|$ is the length of the text, $S$ is the time in seconds spent writing
the text. This time starts directly after the first character has been pressed,
and ends when the last letter has been entered. As such, $1$ is subtracted from
$|T|$, as the time spent to find and press the first character cannot be
accurately determined. However, some typing assessments do not subtract $1$ from
$|T|$. $60$ refers to the number of seconds in a minute and $\frac{1}{5}$
normalizes the metric for the average length of words.

Other metrics also measure speed but they aren't as commonly used as \ac{wpm}.
These include Characters per Minute, Gestures per Second, Adjusted Words per
Minute, and Keystrokes per Second

@@

where $|C|$ is the number of correct characters and $|T|$ is the length of the
text.

@@

\subsubsection{Limitations of the Metrics}
These metrics are all based on the inputted characters by the user. These
metrics do not take into account other aspects of keyboard typing such as
posture, hand and wrist positions, and finger placement. Consequently, these
metrics do not give a full picture of the performance of the person typing and
they only provide a cursory view of how a person types.

@@

This curriculum is self-directed and learners can learn at their own pace. At
its core, the curriculum is designed to be 36-week long with 5-10 minutes of
lessons per day. The lessons in the curriculum follow the three stages of
Motor Learning Theory \parencite{kwt}.

@@

\subsection{Finger and Wrist Kinematics}
The way people move their hands, wrists, and fingers differ between each person.
This can be attributed to the different typing styles each person has. One key
difference between people is the angle of the 5th digit.

However, there are some common movements and positions regardless of typing
style: flexion, or the curving of the fingers, across the fingers, is decreasing
across the hand, with the 2nd digit having the least flexion. This may be due to
the instinct to reduce pronation of the hand, which in turn increases the
distance of the 2nd digit to the keyboard. In addition, some people isolate or
extend one of their thumbs, usually the one not used for pressing a key. This is
also true for some people that do not use their 5th digit during typing
\parencite{baker2007}.

@@

For those that reach their keys by stretching their fingers, there is an
increased probability that the wrists and fingers adapt non-neutral postures.
These include wrist extension, ulnar deviation, and pronation, which may cause
musculoskeletal disorders of the upper extremity (\cite{marklin1999} as cited in
\cite{baker2007})

\section{Finger and Hand Tracking}
Finger and Hand tracking is a method of tracking fingers and hands in 3D space
using motion capture systems or computer vision. This technique allows computers
to perform actions and analyses on the motions and positions of these body
parts.

@@

MediaPipe is an open-source computer vision framework that allows developers to
create a perception pipeline. This perception pipeline is a directed graph of
calculators. Data passes through the graph as packets and a group of packets
constitute a data stream. As the data passes through the pipeline, the
calculators, produce the desired output.

@@

The keyboard will be a 60\% keyboard as shown in
Figure~\ref{fig:metho-setup-keeb}. This will limit the necessary mapping for the
algorithm to the alphanumeric portion of the keyboard. This keyboard choice will
also lessen the area that the camera will need to capture, as this keyboard type
is considerably smaller compared to a full-size keyboard. The keyboard will have
key caps and a case in a light color that will contrast the dark surface that it
will be placed on. This is to improve initial keyboard detection. In addition,
the keyboard layout will also be ANSI, due to the availability and widespread
adoption of the layout in the Philippines.

\subsection{Environment}
The environment the setup would be placed in would be a well lit environment
with a light source beside the camera. This light source will evenly light the
keyboard and the fingers used for typing. The light source to be used will be a
common LED desk lamp rated at 9 Watts with 700 lumens. This light will be white
with a color temperature of 6500k.

In addition, the surface where the keyboard is to be placed on will be solid
green without any variation of color. This is to improve initial keyboard
detection.

@@

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[font=\small,thick]
		\node[draw,
			rounded rectangle,
			minimum width=2.5cm,
			minimum height=1cm] (start) {START};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of start,
			minimum width=3.5cm,
			minimum height=1cm
		] (input) {Input Frame};

		\node[draw,
			below=of input,
			minimum width=3.5cm,
			minimum height=1cm] (gray) {Convert to Grayscale};

		\node[draw,
			below=of gray,
			minimum width=3.5cm,
			minimum height=1cm] (sobel) {Edge Detection using Sobel Filter};

		\node[draw,
			below=of sobel,
			minimum width=3.5cm,
			minimum height=1cm] (contours) {Find Contours (Simple)};

		\node[draw,
			below=of contours,
			minimum width=3.5cm,
			minimum height=1cm] (map) {Overlay Virtual Keyboard Map using Contour Points};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of map,
			minimum width=3.5cm,
			minimum height=1cm
		] (output) {Output Frame};

		\node[draw,
			below=of output,
			rounded rectangle,
			minimum width=2.5cm,
			minimum height=1cm] (end) {END};

@@

\begin{figure}[H]
	\centering
  $\begin{bmatrix}
  +1 & 0 & -1\\
  +2 & 0 & -2\\
  +1 & 0 & -1
  \end{bmatrix}$
	and
  $\begin{bmatrix}
  +1 & +2 & +1\\
  0 & 0 & 0\\
  -1 & -2 & -1
  \end{bmatrix}$
	\caption{Sobel Operator Kernels. Reproduced from \fullcite{sobel2014}}
	\label{fig:metho-algo-key-sobel}
\end{figure}

@@

\subsubsection{Find Contours (Simple)}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. In this algorithm the OpenCV function
\texttt{findContours} will be used to find the contours of the outlines of the
object found using the previous step. This function accepts a contour
approximation method as one of its arguments. Two methods are provided by
OpenCV, \texttt{CHAIN\_APPROX\_NONE} and \texttt{CHAIN\_APPORX\_SIMPLE}. The
former returns all contours in the shape, while the latter removes redundant
points and returns the least amount of points that describes the shape. The
algorithm will use the latter, as only the extreme edges of the keyboard needs
to be detected. This OpenCV function implements the algorithm of
\cite{contours} in their paper ``\citefield{contours}{title}''.

@@

\subsubsection{Overlay Virtual Keyboard Map using Contour Points}
The virtual keyboard map is a rectangular image that contains individual region
of interest (ROI) for each key in a 60\% \ac{ansi} keyboard. Each key has a
corresponding color assigned to it and this color fills the region where this
key is located at. Figure~\ref{fig:metho-algo-key-map} is the virtual keyboard
map.

@@

\subsubsection{Palm Detection Model}
The first model, the Palm Detection Model detects the initial hand locations
using a single shot detector model based on the paper by \cite{ssd}. This model
achieves an average precision of 95.7\% in palm detection
\parencite{mediapipe-hands}.

MediaPipe Hands detects the palms first, instead of the whole hands with one
model because the hands lack high contrast patterns. This reduces the model's
ability to detect the hand with accuracy. In addition, detecting a palm is
simpler compared to detecting hands with articulated fingers since estimating a
bounding box around a rigid objects, i.e.\ a palm, is much simpler. Furthermore,
a palm can be modeled using only square anchors reducing the number of anchors
by a factor of 3--5 \parencite{mediapipe-hands}.

@@

The integration of the algorithm is a two step process. The first step is to get
the mapping of the keyboard using the algorithm shown in
Section~\ref{section:metho-algo-keyboard}.

The second step is a continuous loop where the mapping of the keyboard is used
in conjunction with the finger tracking algorithm chosen in
Section~\ref{section:metho-algo-finger}. The flowchart of the algorithm is shown
in Figure~\ref{fig:metho-algo-integration}

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[font=\small,thick]
		\node[draw,
			rounded rectangle,
			minimum width=2.5cm,
			minimum height=1cm] (start) {START};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below left=of start,
			minimum width=3.5cm,
			minimum height=1cm
		] (map) {Virtual Keyboard Map};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below right=of start,
			minimum width=3.5cm,
			minimum height=1cm
		] (track) {Finger Tracking Frame};

		\coordinate (CENTER) at ($(map)!0.5!(track)$);

		\node[draw,
			diamond,
			aspect=2,
			below = 3cm of CENTER,
			minimum width=2.5cm,
			inner sep=0] (if-packet) {Keypress Packet Received};

		\node[draw,
			below left=of if-packet,
			minimum width=3.5cm,
			minimum height=1cm] (find-roi) {Find ROI of the pressed key in Virtual Keyboard Map};

		\node[draw,
			below=of find-roi,
			minimum width=3.5cm,
			minimum height=1cm] (find-finger) {Find which fingertip is on that ROI};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of find-finger,
			minimum width=3.5cm,
			minimum height=1cm
		] (return) {Return which fingertip};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below = 11cm of CENTER,
			minimum width=3.5cm,
			minimum height=1cm
		] (output) {Output Frame};

		\node[draw,
			below=of output,
			rounded rectangle,
			minimum width=2.5cm,
			minimum height=1cm] (end) {END};

		\draw[-latex] (start) edge (map)
		(start) edge (track)
		(track) edge (if-packet)
		(map) edge (if-packet)
		(find-roi) edge (find-finger)
		(find-finger) edge (return)
		(return) edge (output)
		(output) edge (end);

		\draw[-latex] (if-packet) -| (find-roi)
		node[pos=0.25,fill=white,inner sep=0]{true};

		\draw[-latex] (if-packet) edge node[pos=0.4,fill=white,inner sep=2pt]{No}(output);

@@

\subsubsection{Find ROI of the pressend key in Virtual Keyboard Map}
Since there is a known color-key mapping based on the Virtual Keyboard Map, the
algorithm will find the correct color to search for based on the received
keypress packet. This color is isolated from the Virtual Keyboard Map, and the
region of interest and its coordinates are obtained. The edges of this ROI will
then be obtained by finding its contours. The specific implementation of this is
explained in Section~\ref{section:metho-algo-key-contours}.

@@

The expected output of this is four sets of pairs. Each pair corresponds to an
edge of the region of interest.

\subsubsection{Find which fingertip is on that ROI}
The Finger Tracking Frame will contain the pixel positions of each landmark of
the hand. For this step, the algorithm will find the landmark which is
positioned within the region of interest obtained from the previous step. This
will be done using a series of checks.

Each landmark's coordinates will be compared to the coordinates of the edges of
the ROI. A landmark will be determined as positioned within the ROI if all of the
following conditions are true:
\begin{enumerate}
	\item In the X axis, the landmark's coordinates is greater than one or both of
		the two coordinates found of the left side of the ROI
	\item In the X axis, the landmark's coordinates is less than one or both of
		the two coordinates found of the right side of the ROI
	\item In the Y axis, the landmark's coordinates is greater than one or both of
		the two coordinates found of the top side of the ROI
	\item In the Y axis, the landmark's coordinates is less than one or both of
		the two coordinates found of the bottom side of the ROI
\end{enumerate}

These conditions maximize the total area of the ROI and is not strict about
exact accuracy. In essence, these conditions will create a perfectly rectangular
box that contains the quadrilateral formed by the four points from the previous
step.

\subsubsection{Return which fingertip}
The landmark will then be used to determine which specific finger corresponds to
the keypress. Figure~\ref{fig:metho-algo-integration-landmarks} shows each
possible landmark that may be returned from the previous step. This step will
then return the name of the landmark, up until the first underscore. As an
example, if the landmark found is \texttt{MIDDLE\_FINGER\_TIP}, this step will
return \texttt{MIDDLE} denoting that the middle finger is the finger that
corresponds with the keypress. In addition, the specific hand will also be
returned as one of two strings, \texttt{Left} and \texttt{Right}, since
this information is also bundled together with the landmarks.

@@

\subsection{Typing Test Sequences}
Typing test sequences are strings that will be used in testing the user in their
ability to type. The test sequences that will be used in the trainer will come
from text found in the public domain obtained from Project Gutenberg and the
Internet Archive. Sentences will be isolated from these text and used as test
sequences if they fit a criteria.

@@

There will be a total of 10 test sequences that will be gathered. An example
test sequence is ``What of it, if some old hunks of a sea-captain orders me to
get a broom and sweep down the decks?'' from Moby Dick by \cite{moby-dick}. The
9 other test sequences can be found in the appendix.

\subsection{UX/UI design and Front-end development for the touch typing trainer}
\subsubsection{Design}
This design will be based on other type tests such as Monkeytype
\parencite{bartnik2021}, and Keybr \parencite{keybr}, with additional components
for showing real-time finger-key identification and mapping.
Figure~\ref{fig:ui-test} and \ref{fig:ui-stat} illustrates the two main pages of
the design. Additional screens, such as historical statistics, in-platform
tutorial, and a written guide are shown in the appendix.

>>>>

\documentclass{report}
%%% LT-SKIP-BEGIN

@@

\DeclareAcronym{roi}{
	short=ROI,
	long=Region of Interest
}

\DeclareAcronym{cts}{
	short=CTS,
	long=Carpal Tunnel Syndrome
}

\DeclareAcronym{jwt}{
	short=JWT,
	long=JSON Web Tokens
}

\DeclareAcronym{fpacc}{
	short=FP ACC,
	long=Finger Placement Accuracy
}

\DeclareAcronym{hfpacc}{
	short=HFP ACC,
	long=Historical Finger Placement Accuracy
}

@@

%Tables
\usepackage{booktabs}
\usepackage{supertabular}

@@

%%% LT-SKIP-END

@@

	\textsc{Development of a touch typing trainer\\with an emphasis on finger-key identification }
	\par\vspace{0.5cm}

	\hrulefill{}
	\par\vspace{0.25cm}
	A Special Project Proposal Presented to the\\
	Faculty of the Department of Computer Science,\\
	University of the Philippines Cebu

	\par\vspace{0.25cm}
	In Partial Fulfillment\\
	Of the Requirements for the Degree\\
	Bachelor of Science in Computer Science\\
	\par\vspace{0.25cm}
	\hrulefill{}
	\par\vspace{0.5cm}

@@

	February 2022
	\vfill
	\hspace{0pt}
\end{titlepage}

\newpage

\printacronyms{}
\newpage

\listoffigures
\newpage

@@

\section{Background of the Study}
There are a lot of educational typing tests available that help people learn
touch typing, including Monkeytype, TypeRacer, and Keybr. These typing tests
list out words that are then typed out. The entered keys are then compared to
check if the user has typed the expected letter. At the end of the test, the
time taken is calculated, and certain metrics are given. These metrics include
\ac{wpm} and Accuracy \parencite{bartnik2021}.

@@

Correct finger placement is usually taught at the beginning using a diagram,
with each key being associated with a specific finger. For instance, the letter
Q in a QWERTY layout should be hit using the fifth digit, or the little finger,
of the left hand, and this is shown by coloring the fifth digit and the key Q
with the same color or by placing the letters directly on the fingers
\parencite{dobson2009touch}.

Incorrect finger placement may cause these hand and wrist positions: ulnar
deviation, forearm pronation, and wrist extension \parencite{serina1999}. These
three are hand and wrist positions that are common in all activities, however,
prolonged periods in these positions may cause injuries such as \ac{cts}
\parencite{toosi2015}

@@

Thus, there is a need for automatically detecting which finger is used during
typing. One way to do this is through finger and hand tracking. One solution for
tracking is by using image processing and machine learning. An example of this
is MediaPipe by~\cite{mediapipe}.

MediaPipe allows for various applications for machine learning in the field of
image processing. This includes, hand tracking, pose estimation, object
detection, and others. Another example of a library that allows for hand and
finger tracking is OpenCV by~\cite{opencv}. This is a tool that simplifies
computer vision and image processing. Machine learning can also be used with
OpenCV.\@

@@

\subsection{Specific Objectives}
\begin{enumerate}
	\item Develop an algorithm or technique using computer vision that is capable
	      of identifying which keys are being pressed by which fingers during keyboard
	      typing --- i.e.\ finger-key identification.
	\item Identify the setup and configuration of a single optical camera that can
	      be used to capture images during keyboard typing for real-time finger-key
	      identification using the algorithm identified in 1
	\item Design and develop a touch typing trainer using the algorithm in 1 and
	      camera setup and configuration in 2
	\item Determine the accuracy of the touch typing trainer in finger-key
	      identification
	\item Develop a new keyboard typing metric that uses both the key and
	      finger used to press as parameters
\end{enumerate}

@@

In addition, the keyboard layout will be \ac{ansi}. This layout is described by
the American National Standards Institute \parencite{ansi}. This is the most
common layout in the United States. However, it is also used in numerous
English-speaking countries such as the Philippines, Malaysia, and India
\parencite{apple-layout}.

@@

Capturing of the video to be analyzed by the program would be limited to a
single 720p webcam that is capturing in 30 frames per second. The camera will be
pointed downwards facing the keyboard and the hand.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth]{60.jpg}
	\caption{A 60\% keyboard in \ac{ansi} layout. Reprinted from \fullcite{60}}
	\label{fig:60}
	\centering
\end{figure}

\section{Significance of the Research}
This research is beneficial for all users of physical keyboards. These include a
vast majority of the population as there are a lot of professions that heavily
rely on keyboards. Examples include developers, physicians, educators, and
accountants. By having better ergonomics while typing, wrist injuries can be
prevented, and typing speed may be increased

@@

This research has a direct impact on people that has hand or wrist injuries that
are caused by poor typing habits. By correcting these poor habits, pain from
these injuries will be lessened, and even be prevented from occurring in the
first place. A specific example of this is by reducing ulnar deviation which
affects the nerve that is indicative of \ac{cts} \parencite{toosi2015}.

\chapter{Advanced Review of Related Literature}

@@

Where, $|T|$ is the length of the text, $S$ is the time in seconds spent writing
the text. This time starts directly after the first character has been pressed,
and ends when the last letter has been entered. As such, $1$ is subtracted from
$|T|$, as the time spent to find and press the first character cannot be
accurately determined. However, some typing assessments do not subtract $1$ from
$|T|$. $60$ refers to the number of seconds in a minute and $\frac{1}{5}$
normalizes the metric for the average length of words.

Other metrics also measure speed, but they aren't as commonly used as \ac{wpm}.
These include Characters per Minute, Gestures per Second, Adjusted Words per
Minute, and Keystrokes per Second

@@

Where $|C|$ is the number of correct characters and $|T|$ is the length of the
text.

@@

\subsubsection{Limitations of the Metrics}
These metrics are all based on the inputted characters by the user. These
metrics do not take into account other aspects of keyboard typing such as
posture, hand and wrist positions, and finger placement. Consequently, these
metrics do not give a full picture of the performance of the person typing, and
they only provide a cursory view of how a person types.

@@

This curriculum is self-directed and learners can learn at their own pace. At
its core, the curriculum is designed to be 36-week long with 5--10 minutes of
lessons per day. The lessons in the curriculum follow the three stages of
Motor Learning Theory \parencite{kwt}.

@@

\subsection{Finger and Wrist Kinematics}
The way people move their hands, wrists, and fingers differ between each person.
This can be attributed to the different typing styles each person has. One key
difference between people is the angle of the fifth digit.

However, there are some common movements and positions regardless of typing
style: flexion, or the curving of the fingers, across the fingers, is decreasing
across the hand, with the 2nd digit having the least flexion. This may be due to
the instinct to reduce pronation of the hand, which in turn increases the
distance of the 2nd digit to the keyboard. In addition, some people isolate or
extend one of their thumbs, usually the one not used for pressing a key. This is
also true for some people that do not use their fifth digit during typing
\parencite{baker2007}.

@@

For those that reach their keys by stretching their fingers, there is an
increased probability that the wrists and fingers adapt non-neutral postures.
These include wrist extension, ulnar deviation, and pronation, which may cause
musculoskeletal disorders of the upper extremity (\cite{marklin1999} as cited in
~\cite{baker2007})

\section{Finger and Hand Tracking}
Finger and Hand tracking is a method of tracking fingers and hands in 3D space
using motion capture systems or computer vision. This technique allows computers
to perform actions and analyzes on the motions and positions of these body
parts.

@@

MediaPipe is an open-source computer vision framework that allows developers to
create a perception pipeline. This perception pipeline is a directed graph of
calculators. Data passes through the graph as packets and a group of packets
constitutes a data stream. As the data passes through the pipeline, the
calculators, produce the desired output.

@@

The keyboard will be a 60\% keyboard as shown in
Figure~\ref{fig:metho-setup-keeb}. This will limit the necessary mapping for the
algorithm to the alphanumeric portion of the keyboard. This keyboard choice will
also lessen the area that the camera will need to capture, as this keyboard type
is considerably smaller compared to a full-size keyboard. The keyboard will have
keycaps and a case in a light color that will contrast the dark surface that it
will be placed on. This is to improve initial keyboard detection. In addition,
the keyboard layout will also be ANSI, due to the availability and widespread
adoption of the layout in the Philippines.

\subsection{Environment}
The setup would be placed in would be a well lit environment with a light source
beside the camera. This light source will evenly light the keyboard and the
fingers used for typing. The light source to be used will be a common LED desk
lamp rated at 9 Watts with 700 lumens. This light will be white with a color
temperature of 6500k.

In addition, the surface where the keyboard is to be placed on will be solid
black without any variation of color. This is to improve initial keyboard
detection.

@@

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[font=\small,thick]
		\node[draw,
			rounded rectangle,
			minimum width = 2.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (start) {START};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of start,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (input) {Input Frame};

		\node[draw,
			below=of input,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (gray) {Convert to Grayscale};

		\node[draw,
			below=of gray,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (sobel) {Edge Detection using Sobel Filter};

		\node[draw,
			below=of sobel,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (contours) {Find Contours (Simple)};

		\node[draw,
			below=of contours,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (map) {Overlay Virtual Keyboard Map using Contour Points};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below=of map,
			minimum width=3.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (output) {Output Frame};

		\node[draw,
			below=of output,
			rounded rectangle,
			minimum width=2.5cm,
			minimum height=1cm,
			inner sep = 3mm
		] (end) {END};

@@

\begin{figure}[H]
	\centering
	$\begin{bmatrix}
			+1 & 0 & -1 \\
			+2 & 0 & -2 \\
			+1 & 0 & -1
		\end{bmatrix}$
	and
	$\begin{bmatrix}
			+1 & +2 & +1 \\
			0  & 0  & 0  \\
			-1 & -2 & -1
		\end{bmatrix}$
	\caption{Sobel Operator Kernels. Reproduced from \fullcite{sobel2014}}
	\label{fig:metho-algo-key-sobel}
\end{figure}

@@

\subsubsection{Find Contours (Simple)}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function \texttt{findContours}
will be used to find the contours of the outlines of the object found using the
previous step. This function accepts a contour approximation method as one of
its arguments. Two methods are provided by OpenCV:\@\texttt{CHAIN\_APPROX\_NONE}
and \texttt{CHAIN\_APPROX\_SIMPLE}. The former returns all contours in the
shape, while the latter removes redundant points and returns the least amount of
points that describes the shape. The algorithm will use the latter, as only the
extreme edges of the keyboard needs to be detected. This OpenCV function
implements the algorithm of~\cite{contours} in their paper
\citetitle{contours}.

@@

\subsubsection{Overlay Virtual Keyboard Map using Contour Points}
The virtual keyboard map is a rectangular image that contains individual
\ac{roi} for each key in a 60\% \ac{ansi} keyboard. Each key has a corresponding
color assigned to it and this color fills the region where this key is located
at. Figure~\ref{fig:metho-algo-key-map} is the virtual keyboard map. Appendix
Chapter~\ref{appendix:key-map} is a table that shows the hex color code and key
mapping present in the image map.

@@

\subsubsection{Palm Detection Model}
The first model, the Palm Detection Model detects the initial hand locations
using a single shot detector model based on the paper by~\cite{ssd}. This model
achieves an average precision of 95.7\% in palm detection
\parencite{mediapipe-hands}.

MediaPipe Hands detects the palms first, instead of whole hands with one model
because hands lack high contrast patterns. This reduces the model's ability to
detect hands with accuracy. In addition, detecting a palm is simpler compared
to detecting hands with articulated fingers since estimating a bounding box
around a rigid object, i.e.\ a palm, is much simpler. Furthermore, a palm can be
modeled using only square anchors reducing the number of anchors by a factor of
3--5 \parencite{mediapipe-hands}.

@@

The integration of the algorithm is a two-step process. The first step is to get
the mapping of the keyboard using the algorithm shown in
Section~\ref{section:metho-algo-keyboard}.

The second step runs whenever a key press has been detected. This step uses the
mapping of the keyboard in conjunction with the finger tracking algorithm
selected in Section~\ref{section:metho-algo-finger}. The flowchart of the
algorithm is shown in Figure~\ref{fig:metho-algo-integration}

\begin{figure}[H]
	\centering
	\tikzset{
		rectangle connector/.style={
				-latex,
				to path={(\tikztostart) -- ++(6cm,0pt) \tikztonodes |- (\tikztotarget) },
				pos=0.5
			},
	}
	\begin{tikzpicture}[font=\small,thick]
		\node[draw,
			rounded rectangle,
			minimum width = 2.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (start) {START};

		\node[draw,
			diamond,
			aspect = 2,
			below = of start,
			minimum width = 4cm,
			inner sep = 3mm] (if-packet) {Keypress Detected};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below = 2cm of if-packet,
			minimum width = 3.5cm,
			minimum height = 1cm,
			align = center,
			inner sep = 3mm
		] (map) {Virtual Keyboard Map \\ Finger Tracking Data};

		\node[draw,
			below = of map,
			minimum width = 3.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (find-roi) {Find \ac{roi} of the pressed key in Virtual Keyboard Map};

		\node[draw,
			below=of find-roi,
			minimum width = 3.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (find-finger) {Find which fingertip is on that \ac{roi}};

		\node[draw,
			trapezium,
			trapezium left angle = 65,
			trapezium right angle = 115,
			trapezium stretches,
			below = of find-finger,
			minimum width = 3.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (return) {Return which fingertip};

		\node[draw,
			below = of return,
			rounded rectangle,
			minimum width = 2.5cm,
			minimum height = 1cm,
			inner sep = 3mm
		] (end) {END};

		\draw[-latex] (start) edge (if-packet)
		(map) edge (find-roi)
		(find-roi) edge (find-finger)
		(find-finger) edge (return)
		(return) edge (end);

		\draw[-latex] (if-packet) edge node[pos=0.5, fill=white, inner sep=2] {true} (map);
		\draw[rectangle connector] (if-packet) to node[fill=white, inner sep=2] {false} (end);

@@

\subsubsection{Find \ac{roi} of the pressed key in Virtual Keyboard Map}
Since there is a known color-key mapping based on the Virtual Keyboard Map, the
algorithm will find the correct color to search for based on the received key
press packet. This color is isolated from the Virtual Keyboard Map, and the
\ac{roi} and its coordinates are obtained. The edges of this singular \ac{roi}
will then be obtained by finding its contours. The specific implementation of
this is explained in Section~\ref{section:metho-algo-key-contours}.

@@

The expected output of this is four sets of pairs. Each pair corresponds to an
edge of the single \ac{roi}.

\subsubsection{Find which fingertip is on that \ac{roi}}
The Finger Tracking Data will contain the pixel positions of each landmark of
the hand. For this step, the algorithm will find the landmark which is
positioned within the single \ac{roi} obtained from the previous step.
This will be done using a series of checks.

Each landmark's coordinates will be compared to the coordinates of the edges of
the \ac{roi}. A landmark will be determined as positioned within the \ac{roi} if all the
following conditions are true:
\begin{enumerate}
	\item In the X axis, the landmark's coordinates is greater than one or both of
	      the two coordinates found of the left side of the \ac{roi}
	\item In the X axis, the landmark's coordinates is less than one or both of
	      the two coordinates found of the right side of the \ac{roi}
	\item In the Y axis, the landmark's coordinates is greater than one or both of
	      the two coordinates found of the top side of the \ac{roi}
	\item In the Y axis, the landmark's coordinates is less than one or both of
	      the two coordinates found of the bottom side of the \ac{roi}
\end{enumerate}

These conditions maximize the total area of the \ac{roi} and is not strict about
exact accuracy. In essence, these conditions will create a perfectly rectangular
box that contains the quadrilateral formed by the four points from the previous
step.

\subsubsection{Return which fingertip}
The landmark will then be used to determine which specific finger corresponds to
the key press. Figure~\ref{fig:metho-algo-integration-landmarks} shows each
possible landmark that may be returned from the previous step. This step will
then return the name of the landmark, up until the first underscore. As an
example, if the landmark found is \texttt{MIDDLE\_FINGER\_TIP}, this step will
return \texttt{MIDDLE} denoting that the middle finger is the finger that
corresponds with the key press. In addition, the specific hand will also be
returned as one of two strings, \texttt{Left} and \texttt{Right}, since
this information is also bundled together with the landmarks.

@@

\subsection{Typing Test Sequences}
Typing test sequences are strings that will be used in testing the user in their
ability to type. The test sequences that will be used in the trainer will come
from text found in the public domain obtained from Project Gutenberg and the
Internet Archive. Sentences will be isolated from these text and used as test
sequences if they fit the criteria.

@@

There will be a total of 10 test sequences that will be gathered. An example
test sequence is ``What of it, if some old hunks of a sea-captain orders me to
get a broom and sweep down the decks?'' from Moby Dick by~\cite{moby-dick}. The
9 other test sequences can be found in Appendix
Chapter~\ref{appendix:test-sequences}.

\subsection{UX/UI design and Front-end development for the touch typing trainer}
\subsubsection{Design}
This design will be based on other type tests such as Monkeytype
\parencite{bartnik2021}, and Keybr \parencite{keybr}, with additional components
for showing real-time finger-key identification and mapping.
Figure~\ref{fig:ui-test} and~\ref{fig:ui-stat} illustrates the two main pages of
the design. Additional screens, such as historical statistics, in-platform
tutorial, and a written guide are shown in the appendix.
